Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-07-03 paper/paper.tex
continue revising section 6

First of all, since there is no first-order data types representing model of stacks,
we introduce a configuration class @Config@ as a state-like definition by eliminating
@konts@ and @ndk@ from the definition of state. The type of return value of the abstract
interpreter accordingly becomes @Set[Config]@.
As we are getting close to achieving abstract definitional interpreters, 
the name of @drive_step@ is updated to @aeval@ (for \emph{abstract eval}).

@@

The continuation @Cont@ as a higher-order function takes a configuration, a set of explored
configurations, as well as a meta-continuation as its arguments. While the meta-continuation
only takes a configuration and a set of explored configurations as arguments.
Now, the skeleton of function @aeval@ looks as follows:

@@

Before jumping into the details of refunctionalizing the continuations and constructing
@new_cont@ and @new_mcont@, we have two important observations in the disentangled AAM
where the continuations were represented by data types:
1) When constructing the @new_ndk@, we observe that it explicitly uses @new_frames@ as one
argument of its top @NDCont@ object. The higer-order version of these continuations follows
the same pattern.
2) In the meta-continuation dispatching function @ndcontinue@, the size of the closures set is
decreasing every time while the other parts of meta-continuation are kept. This means in 
the higher-order version, we also need an ``ability'' to update the top frame of meta-continuation
and keep the rest as the same.

@@

Then to effectively construct the meta-continuation @new_mcont@, we define an auxiliary function
@makeMCont@, which takes a list of closures as arguments and returns the meta-continuation. 
Function @makeMCont@ is a recursive function where the recursive invocation happens when 
constructing the meta-continuation based on the rest of closures set. 
Now we can dynamically generate the meta-continuation inside of a meta-continuation.
Meanwhile, the @new_cont@ we defined above is used here, simply because different closures 
actually share the same continuation.
The case where the empty closures set occurs is also interesting. At this case,
the continuation simply applies its meta-continuation @m@, and the meta-continuation is the
outer meta-continuation @mcontinue@. It would be clear if we inductively think the base case 
where the set contains only one closure, then @makeMCont@ is invoked with an empty set, 
therefore the continuation should do nothing. %TODO: rewrite this part.

@@

One last very important thing is about the initial continuation and meta-continuation.
The initial continuation is invoked when one computation path reaches its end,
so we invoke the meta-continuation @m@ to continue exploring other paths.
The initial meta-continuation is invoked when all computation paths reach their ends,
so we terminate with the explored configurations @seen@. They correspond to the 
cases where @konts@ is empty and @konts@ and @ndk@ are both empty, respectively.

@@

In this section, based on the refunctionalized AAM, we present several other transformations 
toward a big-step abstract definitional interpreter:

@@

\textit{Values.}
To properly represent final values, we introduce the case class @VS@ which
contains a set of storable values, a timestamp, and a store.

@@

An instance of @VS@ represents the computational result of one path in the nondeterministic 
evaluation. The reason that we include a store is that there might be a case in which two 
paths have the same values but different accumulated side effects (e.g., memory allocations).
Since the whole computation is potentially nondeterministic, the type of the final result
values is a collection @Set[VS]@. \\

\textit{Caches.}
% why cache? problem of set of states.
Recall that the latent assumption of the work-list algorithm in classical AAM is that 
if we have seen a state @s@, it means we also have seen all the successors of state @s@.
The caching algorithm we adopt here replays the same assumption but in a big-step 
manner: if we have seen the configuration @c@, then it means we also have seen the 
values that are evaluated from the configuration @c@.
Here we will apply the fixed-point caching algorithm as described from
\citeauthor{darais2017abstracting}'s ADI paper \cite{darais2017abstracting}.
The case class @Cache@ and its operations are defined as follow:

@@

The @Cache@ contains two maps @in@ and @out@ which are both mappings from configurations 
to sets of @VS@. The @in@ cache contains mappings from the previous iteration of evaluation, 
and the @out@ cache contains mappings after the current iteration of evaluation. 
Once we have evaluated a term to some values, we update the @out@ cache. In the next 
iteration, we will use the @out@ from the previous iteration as the @in@ cache for the 
current iteration. \\

\textit{Cache-Passing Style.}
We can now transform our interpreter into cache-passing style. We begin by defining the 
return type @Ans@ as a case class which contains a @VS@ set and a cache. @Ans@ also implements 
its own accumulating operation @++@ which defines as union on @Set[VS]@ and join on caches.
The join operation of caches is defined as joining its two store components.

@@

%TODO: co-inductive
Upon entering @aeval@, we first determine whether the @out@ cache contains some values 
for the current configuration and use them immediately (if present).
This corresponds to small-step AAM with a work-list: when we have seen a state,
we can simply discard the state and continue working through the rest of the work-list @todo@.

If, however, we have not hit the @out@ cache, we retrieve the values from the @in@ cache, 
and update the @out@ cache with values from the @in@ cache. This retrieval from the @in@ 
cache may return an empty set of values if there is no such mapping for the configuration 
from the previous iteration.
In reality, we first conservatively assume that all computations can be diverged, and if 
not, we update its values in the cache after evaluation. In terms of partial orders and 
lattices, it is the case that starts the Kleene iteration from the bottom of the lattice.\\

\textit{Nondeterminism Abstraction.}
We observe that in the refunctionalized AAM, @makeMCont@ actually repeats the prologue
part when matching a @let@ expression: they are both preparing a new environment and store 
for the body expression of a closure. %TODO motivation
We present a nondeterministic operator that abstracts this pattern:

@@

Our function @nd@ is a variant implementation of @fold@ in continuation-passing style,
and @f@ also has to be passed as a function that takes a continuation. \\

\textit{Put Them Together.}
Given the skeleton of @aeval@ in cache-and-continuation-passing style, we may now
use the @nd@ operator to fill the details. The first case of match is that @e@
is a @let@ expression:

@@

We next consider the inner application of @nd@, which evaluates over the set of body's 
value/stores @bdvss@. For each @VS@, the timestamp and store in @VS@ will be instantiated to
the next application of @aeval@. The inner application of @aeval@ evaluates @e@ with the
substituted environment and store, then returns a set of values of one computation path 
to its continuation @bdnd@, where the continuation will accumulate the result and 
transfer the evaluation to the next body value.
The meta-continuation of the inner @nd@ is @clsnd@, meaning that we need to transfer 
to the next target closure of @f@. The meta-continuation of the outer @nd@ is just 
@continue@, as what we normally would do when writing CPS interpreters.

@@

% TODO: explain more on how caches work
Finally, the @analyze@ function (as the entrance of the analysis) does a looping iteration
to find the least fixed-point of cache starting from an empty cache.
If the @out@ cache of this iteration is equivalent to its @in@ cache, which means no more 
fact was discovered during this iteration, then we have reached a fixed-point of cache that 
has over-approximated the concrete evaluations, and thus the result can be returned.
Otherwise, we install the latest @out@ cache to the @in@ cache position when initializing 
the next iteration; in the meanwhile, an empty cache will be used as the @out@ cache. 
Note that the initial continuation of calling @aeval@ updates the values for the initial 
configuration in the final cache.
% TODO: summarize this section

>>>>

First of all, as we are getting close to achieving abstract definitional interpreters, 
the name of @drive_step@ is updated to @aeval@ (for \emph{abstract eval}).
Since there will be no first-order data types representing model of stacks,
we introduce a configuration class @Config@ as a state-like definition by eliminating
@konts@ and @ndk@ from the definition of state. The type of return value of the abstract
interpreter accordingly becomes @Set[Config]@.

@@

The continuation @Cont@ as a higher-order function takes a configuration, a set of explored
configurations, as well as a meta-continuation as its arguments. While the meta-continuation
@MCont@ only takes a configuration and a set of explored configurations as arguments.
Now, the skeleton of function @aeval@ looks as follows:

@@

The continuation and meta-continuation are added as arguments to @aeval@.
Before jumping into the details of refunctionalizing the stack structure and constructing
@new_cont@ and @new_mcont@, we have two important observations in the disentangled AAM
where the continuations were represented by data types:
1) When constructing the @new_ndk@, we observe that it explicitly uses @new_frames@ as one
argument of its top @NDCont@ object. The higer-order version of these continuations follows
the same pattern.
2) In the meta-continuation dispatching function @ndcontinue@, the size of the closures set is
decreasing every time while the other parts of meta-continuation are kept. This means in 
the higher-order version, we also need an ``ability'' to update the top frame of meta-continuation
and keep the rest as the same.

@@

Then to effectively construct the meta-continuation @new_mcont@, we define an auxiliary function
@makeMCont@, which takes a list of closures as arguments and returns the meta-continuation. 
Function @makeMCont@ is a recursive function where the recursive invocation happens when 
constructing the meta-continuation based on the rest of closures set. 
Now we can dynamically generate the meta-continuation inside of a meta-continuation.
Meanwhile, the @new_cont@ we defined above is used here, simply because different closures 
actually share the same continuation.
The case where the empty closures set occurs is also interesting. At this case,
the continuation simply applies its meta-continuation @m@, and the meta-continuation of @aeval@
is the outer meta-continuation @mcontinue@. It would be clear if we inductively think the base case 
where the set contains only one closure, then @makeMCont@ is invoked with an empty set, 
therefore the continuation should do nothing. %TODO: rewrite this part.

@@

One last very important thing is about the initial continuation and meta-continuation.
The initial continuation will be invoked when one computation path reaches its end,
so we invoke the meta-continuation @m@ to continue exploring other paths.
The initial meta-continuation will be invoked when all computation paths reach their ends,
so we terminate with the explored configurations @seen@. They correspond to the 
cases where @konts@ is empty and @konts@ and @ndk@ are both empty, respectively.

@@

In this section, based on the refunctionalized AAM, we further present several other 
transformations toward a big-step abstract definitional interpreter:

@@

\subsubsection{Values}
To properly represent final values, we introduce the case class @VS@ which
contains a set of storable values, a timestamp, and a store.

@@

An instance of @VS@ represents the computational result of one path in the nondeterministic 
evaluation. The reason that we include a store is that there might be a case in which two 
paths have the same values but different accumulated side effects (e.g., memory allocations).
Since the whole computation is potentially nondeterministic, the type of the final result
values is a collection @Set[VS]@.

\subsubsection{Caches}
% why cache? problem of set of states.
Recall that the latent assumption of the work-list algorithm in classical AAM is that 
if we have seen a state @s@, it means we also have seen all the successors of state @s@.
The caching algorithm we adopt here replays the same assumption but in a big-step 
manner: if we have seen the configuration @c@, then it means we also have seen the 
values that are evaluated from the configuration @c@.
Here we will apply the fixed-point caching algorithm as described from
\citeauthor{darais2017abstracting}'s ADI paper \cite{darais2017abstracting}.
The case class @Cache@ and its operations are defined as follow:

@@

The @Cache@ contains two maps @in@ and @out@ which are both mappings from configurations 
to sets of @VS@. The @in@ cache contains mappings from the previous iteration of evaluation, 
and the @out@ cache contains mappings after the current iteration of evaluation. 
Once we have evaluated a term to some values, we update the @out@ cache. In the next 
iteration, we will use the @out@ cache from the previous iteration as the @in@ cache.

\subsubsection{Cache-Passing Style}
%motivation
We can now transform our interpreter into cache-passing style. We begin by defining the 
return type @Ans@ as a case class which contains a @VS@ set and a cache. @Ans@ also implements 
its own accumulating operation @++@ which defines as a union of @Set[VS]@s and a join of caches.
The join operation of caches is defined as joining its two store components.

@@

%TODO: co-inductive
Upon entering @aeval@, we first determine whether the @out@ cache contains some values 
for the current configuration and use them immediately (if present) by calling @continue@.
This corresponds to small-step AAM with a work-list: when we have seen a state,
we can simply discard the state and continue working through the rest of the work-list.

If, however, we have not hit the @out@ cache, we retrieve the values from the @in@ cache, 
and update the @out@ cache with values from the @in@ cache. This retrieval from the @in@ 
cache may return an empty set of values if there is no such mapping for the configuration 
from the last iteration.
In reality, we first conservatively assume that all computations can be diverged, and if 
not, we update its values in the cache after the evaluation. In terms of partial orders and 
lattices, it is the case that starts the Kleene iteration from the bottom of the lattice.

\subsubsection{Nondeterminism Abstraction}
We observe that in the refunctionalized AAM, @makeMCont@ actually repeats the prologue
part when matching a @let@ expression: they are both preparing a new environment and store 
for the body expression of a closure. %TODO motivation
We present a nondeterministic operator that abstracts this pattern:

@@

Our function @nd@ is a variant implementation of @fold@ in continuation-passing style,
and @f@ also has to be passed as a function that takes a continuation.

\subsubsection{Put Them Together}
Given the skeleton of @aeval@ in cache-and-continuation-passing style, we may now
use the @nd@ operator to fill the details. The first case of match is that @e@
is a @let@ expression:

@@

We next consider the inner application of @nd@, which evaluates over the set of body's 
value/stores @bdvss@. For each @VS@, the time-stamp and store in @VS@ will be instantiated to
the next application of @aeval@. The inner application of @aeval@ evaluates @e@ with the
substituted environment and store, then returns a set of values of one computation path 
to its continuation @bdnd@, where the continuation will accumulate the result and 
transfer the evaluation to the next body value.
The meta-continuation of the inner @nd@ is @clsnd@, meaning that we need to transfer 
to the next target closure of @f@. The meta-continuation of the outer @nd@ is just 
@continue@, as what we normally would do when writing CPS interpreters.

@@

% TODO: explain more on how caches work
Finally, the @analyze@ function (as the entrance of the analysis) does a looping iteration
to find the least fixed-point of cache starting from an empty cache.
If the @out@ cache of this iteration is equivalent to its @in@ cache, which means no more 
fact was discovered during this iteration, then we have reached a fixed-point of cache that 
has over-approximated the concrete evaluations, and thus the result can be returned.
Otherwise, we install the latest @out@ cache to the @in@ cache position when initializing 
the next iteration; in the meanwhile, an empty cache will be used as the @out@ cache. 
Note that the initial continuation of calling @aeval@ updates the values for the initial 
configuration in the final cache.

% TODO: summarize this section
